A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-06-17 paper.Rmd
finished draft of data documenting :tada:

**In this paper we focus on the practical side of sharing data for the purpose of reproducibility.** Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of idealitstic ideas such as the FAIR data principles [@Wilkinson2016] since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

@@

There are many features to include with data alongside publications. However, not all of these are needed for every dataset. This section discusses how to think about the degree of documentation to include. To frame discussion around the challenges of data documenting, we can think of how an individual dataset falls on two features: "Effort to prepare", and "Ease of understanding" in Figure \@ref(fig:fig-effort-understanding).

```{r fig-effort-understanding, fig.cap = "There is a big difference in the effort to prepare data, and how easy it is to understand - look at the difference between most datasets, and something like a Randomized Control Trial (RCT).", out.width = "90%"}

@@

dat_prep <- tibble(prep = c(15,20,25,90, 96, 98),
               understanding = c(15,20,95,90,96, 80),
                  label = c("Data Dump",
                            "Most studies",
                            "Sensor Data",
                            "50 hectare",
                            "RCT",
                            "Simulation Study"
                            ))

@@

```{r data-sharing-workflow, fig.cap = "The steps required compared to effort required for data sharing for reproducibility.", out.width = "90%"}
knitr::include_graphics("figures/data_sharing_workflow.png")
```

To decide the extent of data documentation to include should be based on the impact of the data. The more impactful the data, the more features to include.

To improve data impact it needs more time to prepare so it is easier to understand. Accordingly, data with higher potential for impact and reuse should be made as easy to understand as possible. Therefore data with high potential impact requires more time and effort to prepare.

High impact data should be easy to understand and use. Impact is hard to measure as a general measure, and may vary from field to field. An example is given below.

For example, data from medical randomized control trials (RCTs) on cancer treatment can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on simple demographics can have low impact. So it might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means effort put into preparing data documentation for a small survey to be made easy to use should be kept relatively simple and not over complicated. Likewise, data that can be created via simulation from open source software could arguably not be archived - since it can be generated from scratch with code; a reproducible process with likely very low cost in person time to create.

Figure \@ref(fig:data-sharing-workflow) shows the different degrees or steps of this.

Tradeoffs may change over time as the cost of making them easy to prepare and understand change. If it becomes technologically cheaper and easier to document data more rigorously in the same time as it takes to do poorly now, then there is no reason why more people cannot do this, even when the benefits are not immediately apparent. So if something might appear to have low impact, it should still be made easy to understand, if it doesn't take long to prepare that documentation.

Creating good documentation has the same challenge as doing good writing: it takes time, and it can be hard to know when you are done. Thinking about the potential impact of should help guide the researcher in how much they should document their data.

@@

It is not possible to summarise the contents of the data, but briefly, the SDSS data includes photometric and spectroscopic information obtained from a 2.5m telescope at Apache Point Observatory in New Mexico, operating since 2000,s producing over 200Gb of data every day [@sdss-one-paper; @sdss-four-paper; @sdss-website]. . This data has been used to produce the most detailed, three dimensionsal maps of the universe ever made.

@@

@@

There are many features to consider when documenting, sharing, and publishing data. What format to store the data in? How do you describe the variables, observations, and collection methods? Do you include raw data? What about data used in analysis? What about metadata for the variables? And then there's lincenses, and long term storage. It can be hard to know where to start, but together they all help make the data easier to understand. These features are not clearly described in the scientific literature in the context of including them alongside research.

>>>>

**In this paper we focus on the practical side of sharing data for the purpose of reproducibility.** Our goal is to describe various methods in which an analyst can share their data with minimal friction. We steer clear of idealistic ideas such as the FAIR data principles [@Wilkinson2016] since they still do not help a researcher share their data. We also skip the discussion around citation and credit because data citations are still poorly tracked and there is no standardized metric or a h-index equivalent for data as of this writing.

@@

There are many features to include with data alongside publications. However, not all of these are needed for every dataset. Working out which are needed is a challenge that is not discussed in the literature. This section discusses a framework to use decide how much they should document their data. To frame discussion around the challenges of data documenting, we can think of how an individual dataset falls on two features: "Effort to prepare", and "Ease of understanding" in Figure \@ref(fig:effort-understanding). The ideal space to be in the graph would be the top left hand corner. But what we notice is that taking more effort to prepare data means that the data is easier to understand.

```{r effort-understanding, fig.cap = "There is a big difference in the effort to prepare data, and how easy it is to understand - look at the difference between most datasets, and something like a Randomized Control Trial (RCT).", out.width = "90%"}

@@

dat_prep <- tibble(prep = c(15,20,55,90, 96, 40),
               understanding = c(15,20,95,90,96, 91),
                  label = c("Data Dump",
                            "Most studies",
                            "Sensor Data",
                            "50 hectare",
                            "RCT",
                            "Simulation Study"
                            ))

@@

Data with higher potential impact and reuse should be made as easy to understand as possible; but it also requires more time and effort to prepare. Impact is hard to measure, and varies from field to field, but as an example, take some data from medical randomized control trials (RCTs) on cancer treatment. These can have high impact, so requires a lot of time and effort document. Comparatively, a small survey on a few demographics can have low impact. Placing these on the graph above, we see they might not have a worthwhile tradeoff for ease of understanding and ease of preparation. This means the effort put into preparing data documentation for a small survey should be kept relatively simple, not over complicated. Likewise, data that can be created via simulation from open source software could arguably not be shared since it can be generated from scratch with code; a reproducible process that requires computer time, not person time to create.

Deciding how much data documentation to include should be based on the data's impact. The more impactful the data, the more documentation features to include. Figure \@ref(fig:data-sharing-workflow) shows the practical types of steps that can be taken against the effort required.

```{r data-sharing-workflow, fig.cap = "The steps required compared to effort required for data sharing for reproducibility.", out.width = "90%"}
knitr::include_graphics(here::here("figures",
                                   "data_sharing_workflow.png"))
```

Documentation challenges may evolve over time as the cost of making them easy to prepare and understand change. For example, new technologically may automate rigorous data documentation, so thorough documentation can take the same time as it takes to do poorly now. In this case, there would be no reason why more people cannot do this, even when the benefits are not immediately apparent. So if something might appear to have low impact, it should still be made easy to understand, if it doesn't take long to prepare that documentation.

Creating good documentation has similar challenges to good writing: it takes time, and it can be hard to know when you are done. Thinking about two features: 1) the impact of data, and 2) the current effort to do each step, provides guidelines for the researcher to decide how much documentation they should provide for their data.

@@

It is not possible to summarise the contents of the data, but briefly, the SDSS data includes photometric and spectroscopic information obtained from a 2.5m telescope at Apache Point Observatory in New Mexico, operating since 2000,s producing over 200Gb of data every day [@sdss-one-paper; @sdss-four-paper; @sdss-website]. . This data has been used to produce the most detailed, three dimensional maps of the universe ever made.

@@

To improve data impact it needs more time to prepare so it is easier to understand.

@@

There are many features to consider when documenting, sharing, and publishing data. What format to store the data in? How do you describe the variables, observations, and collection methods? Do you include raw data? What about data used in analysis? What about metadata for the variables? And then there's licenses, and long term storage. It can be hard to know where to start, but together they all help make the data easier to understand. These features are not clearly described in the scientific literature in the context of including them alongside research.
